---
title: "statistical_analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Load libraries
```{r}
library(tidyverse)
library(caret)
library(glmnet)
library(pROC)
library(cli)
library(lightgbm)
setwd("/data/projects/osteoporosis_metabolomics/johan/")
source("~/osteoporosis_metabolomics/utils.R")

```


# Read and combined positive and negative ionization data


```{r}
ms1 <- read_rds("/data/projects/osteoporosis_metabolomics/johan/ml_preprocessed_msobject.rds")
ms2 <- read_rds("/data/projects/osteoporosis_metabolomics/johan/ml_preprocessed_msobject_neg.rds")

ms_pos <- ms1$i %>% bind_cols(ms1$v2)
ms_neg <- ms2$i %>% bind_cols(ms2$v2) %>% filter(rowid != 257) %>% mutate(rowid = row_number())
ms_pos <- ms_pos %>% arrange(match(ms_neg$sample_id, ms_pos$sample_id))
```

##  model

## Visualizing replicate feature correlations before and after preprocessing

```{r}
if (!file.exists("~/osteoporosis_metabolomics/models/ms_fits_simple_glmnet.rds") |
    !file.exists("~/osteoporosis_metabolomics/models/ms_fits_simple_ranger.rds") |
    !file.exists("~/osteoporosis_metabolomics/models/ms_fits_slope_glmnet.rds" ) |
    !file.exists("~/osteoporosis_metabolomics/models/ms_fits_slope_ranger.rds" )){
  ms_fits_simple_glmnet <- 
    list(ms_neg, ms_pos) %>% 
    map(fit, objective = "simple", cv_method = "sample_origin", method = "glmnet", tuneLength = 5)
  
  ms_fits_slope_glmnet <- 
    list(ms_neg, ms_pos) %>% 
    map(fit, objective = "slope", cv_method = "loocv", method = "glmnet", tuneLength = 5)
  
  ms_fits_simple_ranger <- 
    list(ms_neg, ms_pos) %>% 
    map(fit, objective = "simple", cv_method = "sample_origin", method = "ranger", tuneLength = 5)
  
  ms_fits_slope_ranger <- 
    list(ms_neg, ms_pos) %>% 
    map(fit, objective = "slope", cv_method = "loocv", method = "ranger", tuneLength = 5)
} else {
    ms_fits_simple_glmnet <- readr::read_rds("~/osteoporosis_metabolomics/models/ms_fits_simple_glmnet.rds")
    ms_fits_slope_glmnet  <- readr::read_rds("~/osteoporosis_metabolomics/models/ms_fits_slope_glmnet.rds")
    ms_fits_simple_ranger <- readr::read_rds("~/osteoporosis_metabolomics/models/ms_fits_simple_ranger.rds" )
    ms_fits_slope_ranger  <- readr::read_rds("~/osteoporosis_metabolomics/models/ms_fits_slope_ranger.rds" )
}

ms_preds_glmnet <- 
  c(ms_fits_simple_glmnet, ms_fits_slope_glmnet) %>% 
  get_best_glmnet_preds %>% 
  set_names(c("simple_neg", "simple_pos", "slope_neg", "slope_pos"))

ms_preds_ranger <- 
  c(ms_fits_simple_ranger, ms_fits_slope_ranger) %>% 
  get_best_ranger_preds %>% 
  set_names(c("simple_neg", "simple_pos", "slope_neg", "slope_pos"))
```

# Model performance
```{r}
ms_preds_glmnet %>% 
  map(~{confusionMatrix(.x$pred, .x$obs, mode = "prec_recall")})

ms_preds_ranger %>% 
  map(~{confusionMatrix(.x$pred, .x$obs, mode = "prec_recall")})

ms_preds_glmnet %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]]) 

ms_preds_ranger %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]]) 

```


```{r}

if (!file.exists("./ms_simple_permuted_loocv.rds")){
  ms_simple_permuted <- 
    list(ms_neg) %>% 
    map(fit, permute = T, permutations = 10, method = "glmnet", objective = "simple")

  readr::write_rds(ms_simple_permuted, "./ms_simple_permuted_loocv.rds")
} else{
  ms_simple_permuted <- readr::read_rds("./ms_simple_permuted_loocv.rds")
}
```


```{r}
prediction_both_models <- ms_preds_glmnet %>% 
  set_names(c("Negative_Latest Sample", "Positive_Latest Sample", "Negative_Slope", "Positive_Slope"))

permutation <- readr::read_rds("./ms_simple_permuted_loocv.rds")

glmnet_result <- model_result(prediction_both_models = prediction_both_models, permutation = permutation)

ggsave(filename = "~/osteoporosis_metabolomics/export/figure2ab.pdf",
       plot=glmnet_result,
       bg = "white", width = 183, height = 100, units = "mm")


prediction_both_models <- ms_preds_ranger %>% 
  set_names(c("Negative_Latest Sample", "Positive_Latest Sample", 
              "Negative_Slope", "Positive_Slope"))


ranger_result <- model_result(prediction_both_models = prediction_both_models, permutation = permutation)

```


save the plotting data for extended data
```{r}
prediction_both_models <- ms_preds_glmnet

names(prediction_both_models) <- c("Latest_Negative", "Latest_Positive", "Slope_Negative", "Slope_Positive")

ms_permuted <- readr::read_rds("./ms_slope_permuted_loocv.rds")

HA_ROC <- 
  prediction_both_models %>% 
  map(~pROC::roc(.x$obs, .x$case, )) %>% 
  map(~{pROC:::get.coords.for.ggplot(.x) %>% as_tibble()}) %>% 
  bind_rows(.id = "dataset") %>% 
  mutate(Ionization = strsplit(dataset, "_") %>% map_chr(2),
         method = strsplit(dataset, "_") %>% map_chr(1))

HA_ROC %>% 
  write_csv("~/osteoporosis_metabolomics/export/HA_ROC.csv")


H0_ROC <- 
  ms_permuted[[1]] %>% 
      map(~.x$pred %>% filter(alpha==.x$bestTune$alpha, lambda==.x$bestTune$lambda)) %>% 
      map(~arrange(.x, rowIndex)) %>% 
      map(~pROC::roc(.x$obs, .x$case, )) %>% 
      map_dfr(pROC:::get.coords.for.ggplot) %>% 
      as_tibble() %>% 
      group_by(`1-specificity`) %>% 
      summarise(
        sens_mean = mean(sensitivity),
        upper = sens_mean+1.96*sqrt((sens_mean*(1-sens_mean))/length(ms_permuted[[1]])),
        lower = sens_mean-1.96*sqrt((sens_mean*(1-sens_mean))/length(ms_permuted[[1]])),
        sens_lower = quantile(sensitivity, 0.05),
        sensitivity = quantile(sensitivity, 0.95))

H0_ROC %>% 
  write_csv("~/osteoporosis_metabolomics/export/H0_ROC.csv")

```


# Technical validation
```{r}

ms_slope_fits_batch_val_glmnet <- 
  list(ms_neg, ms_pos)  %>% 
  map(fit, objective = "slope", cv_method = "batch", method = "glmnet")

ms_slope_fits_batch_val_ranger <- 
  list(ms_neg, ms_pos)  %>% 
  map(fit, objective = "slope", cv_method = "batch", method = "ranger")

ms_slope_permute_batch_val_glmnet <- 
  list(ms_neg, ms_pos)  %>% 
  map(fit, objective = "slope", permute = T, permutations = 100, cv_method = "batch", method = "glmnet")


ms_slope_fits_batch_val_glmnet %>% 
  get_best_glmnet_preds() %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]]) 

ms_slope_fits_batch_val_ranger %>% 
  get_best_ranger_preds() %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]]) 
  
```

```{r}
models <- c(ms_slope_fits_batch_val_glmnet %>% get_best_glmnet_preds(),
            ms_slope_fits_batch_val_ranger %>% get_best_ranger_preds()) %>% 
  set_names(c("Negative_Elastic Net", "Positive_Elastic Net",
              "Negative_Random Forest", "Positive_Random Forest")
              )

hold_one_batch_out <- 
  model_result(models, ms_slope_permute_batch_val_glmnet)

ggsave(hold_one_batch_out, filename = "./figures/supplementary_figure_lobathco.pdf", width = 180, height = 60, units = "mm", bg="white")

```


# Jackknife analysis

* Robustness of the features
* Significance of the models (glmnet slope)

```{r}
longitudinal_samples <- 
  ms_neg %>% 
  filter(!is.na(group)) %>% 
  group_by(group) %>% 
  mutate(n=n()) %>% 
  ungroup() %>% 
  filter(n>4) %>% 
  select(-n) %>% 
  mutate(id =  paste0(group, status)) %>% 
  pull(id) %>% 
  unique()

if (!file.exists("~/osteoporosis_metabolomics/jackknifed_models.rdata")){

    
  make_ms_slope_fits <- 
    function(exclusion_sample, 
             exclusion_variable = "none",
             method = "glmnet") list(ms_neg, ms_pos) %>%
    map(fit_slope, 
        cv_method = "loocv", 
        method = "glmnet",
        exclude_sample = exclusion_sample,
        exclude_variable = exclusion_variable)
  
  
  jackknife_slope_models <- 
    longitudinal_samples %>% 
    map(make_ms_slope_fits, .progress=TRUE)
  
  readr::write_rds(file = "~/osteoporosis_metabolomics/jackknifed_models.rdata", jackknife_slope_models)
} else {
  jackknife_slope_models <- readr::read_rds("~/osteoporosis_metabolomics/jackknifed_models.rdata")
}


```

### Finding robustness of feature selection

```{r}
neg_jk_models <- jackknife_slope_models %>% map(1)

feature_importance_neg <- 
  neg_jk_models %>% 
  map(caret::varImp) %>% 
  map(~ as_tibble(.x[[1]], rownames = "feature")) %>% 
  bind_rows(.id = "index")%>% 
  mutate(sample = longitudinal_samples[as.integer(index)])

features_selected_neg <- 
  feature_importance_neg %>% 
  group_by(feature) %>% 
  summarise(feature_mean = mean(Overall)) %>% 
  arrange(-feature_mean)


pos_jk_models <- jackknife_slope_models %>% map(2)


feature_importance_pos <- 
  pos_jk_models %>% 
  map(caret::varImp) %>% 
  map(~ as_tibble(.x[[1]], rownames = "feature")) %>% 
  bind_rows(.id = "index") %>% 
  mutate(sample = longitudinal_samples[as.integer(index)])


features_selected_pos <- 
  feature_importance_pos %>% 
  group_by(feature) %>% 
  summarise(feature_mean = mean(Overall)) %>% 
  arrange(-feature_mean)


```

```{r}
feature_importance_neg %>% 
  group_by(feature) %>% 
  summarize(Overall = mean(Overall)) %>% 
  arrange(-Overall) %>% 
  write_csv("~/osteoporosis_metabolomics/export/feature_importance_neg.csv")

feature_importance_pos %>% 
  group_by(feature) %>% 
  summarize(Overall = mean(Overall)) %>% 
  arrange(-Overall) %>% 
  write_csv("~/osteoporosis_metabolomics/export/feature_importance_pos.csv")
```



### Predict hold-out samples on Random Forest (suggested by reviewer - not included in study)
```{r}

rf_ms_slope_fits <- 
  function(
    ms,
    exclude_samples = "none",
    exclude_features = "none",
    method = "glmnet"){
    
    list(ms) %>%
      map(fit, 
      cv_method = "loocv", 
      objective = "slope",
      method = method,
      exclude_samples = exclude_samples,
      exclude_features = exclude_features)
  }


if (!file.exists("~/osteoporosis_metabolomics/rf_neg_feature_selected.rdata")){
  neg_revision_analysis <-
  as.list(longitudinal_samples) %>%
  map(~feature_importance_neg %>% filter(sample %in% .x)) %>%
  map(filter, Overall == 0) %>%
  map(~{rf_ms_slope_fits(
    ms = ms_neg,
    exclude_samples = .x$sample[1], 
    exclude_features= .x$feature, 
    method = "ranger")},
    .progress = TRUE)

readr::write_rds(file = "~/osteoporosis_metabolomics/rf_neg_feature_selected.rdata", neg_revision_analysis)

}else{
  neg_revision_analysis <- readr::read_rds("~/osteoporosis_metabolomics/rf_neg_feature_selected.rdata")
}

if (!file.exists()){
  pos_revision_analysis <- 
  as.list(longitudinal_samples) %>% 
  map(~feature_importance_pos %>% filter(sample %in% .x)) %>% 
  map(filter, Overall == 0) %>% 
  map(~{rf_ms_slope_fits(
    ms = ms_pos,
    exclusion_sample = .x$sample[1], 
    exclusion_variable=.x$feature, 
    method = "ranger")},
    .progress = TRUE)

 readr::write_rds(file = "~/osteoporosis_metabolomics/rf_pos_feature_selected.rdata", pos_revision_analysis)

} else {
 pos_revision_analysis <- readr::read_rds("~/osteoporosis_metabolomics/rf_pos_feature_selected.rdata") 
}
```


```{r}
ms <- ms_neg
rowinfo <- ms %>% select(!starts_with("M"))
values <- ms %>% select(starts_with("M"), -MCS)
  
tmp1 <- 
  rowinfo %>% 
  filter(!is.na(group)) %>% 
  group_by(group) %>% 
  mutate(n=n()) %>% 
  ungroup() %>% 
  filter(n>4) %>% 
  select(-n)

tmp2 <- values[tmp1$rowid,]
tmp2 <- tmp2[,tmp2 %>% map_lgl(~sum(is.na(.x))==0)]

try2 <- 
  tmp1 %>% 
  select(group, time_point = age_at_donation, rowid, status, batch) %>% 
  bind_cols(tmp2) %>%
  select(-rowid) %>% 
  pivot_longer(cols = -c(group, time_point, status, batch)) %>%
  group_by(group, status, name) %>% 
  mutate(value = scale(value)) %>% 
  summarise(
    batch = batch[1],
    metric = sum((time_point-mean(time_point, na.rm = T))*(value-mean(value, na.rm = T)))/(sum((time_point-mean(time_point, na.rm = T))^2, na.rm = T))
  ) %>% ungroup() %>% 
  pivot_wider(names_from = name, values_from = metric) %>% 
  ungroup()


names(neg_revision_analysis) <- longitudinal_samples

neg_hold_out_predictions_random_forest <- 
  longitudinal_samples %>% 
  map(~{
    x <- try2 %>% filter(paste0(group, status)==.x) %>% select(starts_with("M"))
    x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0] %>%
      select(any_of(colnames(neg_revision_analysis[[.x]][[1]]$trainingData)))
    predict(neg_revision_analysis[[.x]][[1]], newdata = x, type = "prob")[1,]
  }) %>% 
  unlist() #%>% as.factor()

neg_revision_analysis %>% 
  map(~.x[[1]]$pred %>% as_tibble() %>% filter(mtry == .x[[1]]$bestTune$mtry, splitrule == .x[[1]]$bestTune$splitrule)) %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]]) %>% mean()


# Performance when using a two-step approach (first feature selection, then re-training on selected features)
pROC::roc(as.factor(names(neg_hold_out_predictions_random_forest)), as.numeric(neg_hold_out_predictions_random_forest))

```


### Positive ESI two step performance

```{r}
ms <- ms_pos
rowinfo <- ms %>% select(!starts_with("M"))
values <- ms %>% select(starts_with("M"), -MCS)
  
tmp1 <- 
  rowinfo %>% 
  filter(!is.na(group)) %>% 
  group_by(group) %>% 
  mutate(n=n()) %>% 
  ungroup() %>% 
  filter(n>4) %>% 
  select(-n)

tmp2 <- values[tmp1$rowid,]
tmp2 <- tmp2[,tmp2 %>% map_lgl(~sum(is.na(.x))==0)]

try2 <- 
  tmp1 %>% 
  select(group, time_point = age_at_donation, rowid, status, batch) %>% 
  bind_cols(tmp2) %>%
  select(-rowid) %>% 
  pivot_longer(cols = -c(group, time_point, status, batch)) %>%
  group_by(group, status, name) %>% 
  mutate(value = scale(value)) %>% 
  summarise(
    batch = batch[1],
    metric = sum((time_point-mean(time_point, na.rm = T))*(value-mean(value, na.rm = T)))/(sum((time_point-mean(time_point, na.rm = T))^2, na.rm = T))
  ) %>% ungroup() %>% 
  pivot_wider(names_from = name, values_from = metric) %>% 
  ungroup()


names(pos_revision_analysis) <- longitudinal_samples

predict(pos_revision_analysis[["28ctrl"]][[1]], newdata = x)

pos_hold_out_predictions_random_forest <- 
  longitudinal_samples %>% 
  map(~{
    x <- try2 %>% filter(paste0(group, status)==.x) %>% select(starts_with("M"))
    x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0] %>%
      select(any_of(colnames(pos_revision_analysis[[.x]][[1]]$trainingData)))
    predict(pos_revision_analysis[[.x]][[1]], newdata = x)
  }) %>% 
  unlist() %>% 
  as.factor()

caret::confusionMatrix(pos_hold_out_predictions_random_forest, as.factor(gsub('[[:digit:]]+', '', longitudinal_samples)))

```




# Get p-values for jackknife vs permutation
```{r}
ms_slope_permuted <- readr::read_rds("~/osteoporosis_metabolomics/models/ms_slope_permuted_loocv.rds")

neg_perm_auc <- 
  c(ms_slope_permuted[[1]],ms_slope_permuted[[2]])  %>% 
  get_best_glmnet_preds() %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]])

ha_neg <- 
  jackknife_slope_models %>%
  map(1) %>% 
  get_best_glmnet_preds() %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]])

ha_pos <- 
  jackknife_slope_models %>%
  map(2) %>% 
  get_best_glmnet_preds() %>% 
  map_dbl(~pROC::roc(.x$obs, .x$case)$auc[[1]])

sam <- sample(1:length(neg_perm_auc), size=78)
mean(neg_perm_auc[sam])-1.96*(sd(neg_perm_auc[sam])/sqrt(78))
mean(neg_perm_auc[sam])+1.96*(sd(neg_perm_auc[sam])/sqrt(78))

mean(ha_neg)-1.96*(sd(ha_neg)/sqrt(78))
mean(ha_neg)+1.96*(sd(ha_neg)/sqrt(78))
mean(ha_neg)

mean(ha_pos)-1.96*(sd(ha_pos)/sqrt(78))
mean(ha_pos)+1.96*(sd(ha_pos)/sqrt(78))
mean(ha_pos)

t.test(x=neg_perm_auc[sam], y=ha_pos)
t.test(x=neg_perm_auc[sam], y=ha_neg)

```


# Make slope PCA plots

```{r}
  
################### Slopes #################
slopes <- 
  list("neg"=ms_neg, "pos"=ms_pos) %>% 
  map(make_slopes)


source("/data/projects/osteoporosis_metabolomics/johan/scripts/markdown/preprocessing_utils.R")

pca_out_neg <- plot_pca(slopes[[1]]$values, slopes[[1]]$rowinfo, color_labels = c("status", "batch"), palette = c("Set1", "Set1"))

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure1_pca_slopes_neg.pdf",
       plot=pca_out_neg,
       bg = "white", width = 183, height = 100, units = "mm")


pca_out_pos <- plot_pca(slopes[[2]]$values, slopes[[2]]$rowinfo, color_labels = c("status", "batch"), palette = c("Set1", "Set1"))
ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure1_pca_slopes_pos.pdf",
       plot=pca_out_pos,
       bg = "white", width = 183, height = 100, units = "mm")


################### all samples ###################
x <- ms_neg %>% select(starts_with("M"))
x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0]
pca_out_neg_simple <- plot_pca(x, 
                        ms_neg %>% select(-starts_with("M")), color_labels = c("status", "batch"), palette = c("Set1", "Set1"))

r    <- prcomp(x = x, retx = TRUE, center = T, scale = T, rank. = 4*2)
r$x %>% 
  as_tibble() %>% 
  mutate(status = ms_neg$status) %>% 
  write_csv("~/osteoporosis_metabolomics/export/extended_data_plot1_PCA_neg.csv")

pca_out_neg_simple$data

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure1_pca_simple_neg.pdf",
       plot=pca_out_neg_simple,
       bg = "white", width = 183, height = 100, units = "mm")



x <- ms_pos %>% select(starts_with("M"))
x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0]
pca_out_pos_simple <- plot_pca(x, 
                        ms_pos %>% select(-starts_with("M")), color_labels = c("status", "batch"), palette = c("Set1", "Set1"))

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure1_pca_simple_pos.pdf",
       plot=pca_out_pos_simple,
       bg = "white", width = 183, height = 100, units = "mm")

```



# boxplots

```{r}

pos_features <- 
  c("M328.032T305.074", 
    "M297.107T53.58", 
    "M490.149T372.595", 
    "M586.43T746.919", 
    "M326.123T126.859")

neg_features <- 
  c("M515.284T470.37",
    "M320.993T304.505",
    "M169.087T413.776",
    "M227.983T37.03",
    "M363.18T494.785")

features <- list(neg_features, pos_features)


boxplot_data <- 
  slopes[[1]]$values %>% 
    select(all_of(neg_features))%>%
    bind_cols(slopes[[1]]$rowinfo) %>% 
    select(-batch) %>% 
    pivot_longer(cols = starts_with("M")) %>% 
  mutate(ionization = "Negative") %>% 
  bind_rows(
slopes[[2]]$values %>% 
    select(all_of(pos_features))%>%
    bind_cols(slopes[[2]]$rowinfo) %>% 
    select(-batch) %>% 
    pivot_longer(cols = starts_with("M")) %>% 
  mutate(ionization = "Positive")
  ) %>% 
  mutate(name = factor(name, levels = c(neg_features, pos_features))) 

boxplot_data <- 
  boxplot_data %>% 
  mutate(
    Feature = case_when(
      grepl("M328.032T305.074", name)~"Hippuric Acid",
      grepl("M297.107T53.58",   name)~"Tyrosyl-aspartate",
      grepl("M490.149T372.595", name)~"M490.149T372.595",
      grepl("M586.43T746.919",  name)~"M586.43T746.919",
      grepl("M326.123T126.859", name)~"N-fructosyl tyrosine",
      grepl("M515.284T470.37",  name)~"Dodecanedioic Acid",
      grepl("M320.993T304.505", name)~"Hippuric Acid",
      grepl("M169.087T413.776", name)~"Azelaic Acid",
      grepl("M227.983T37.03",   name)~"M227.983T37.03",
      grepl("M363.18T494.785",  name)~"M363.18T494.785"
    )
  ) %>% select(-name)

boxplot_data %>% 
  write_csv("~/osteoporosis_metabolomics/export/extended_data_plot2_boxplot.csv")

boxplot_data %>% 
  ggplot(aes(x=Feature, y=value, fill = status))+
  geom_hline(yintercept = 0, linetype=21)+
  geom_boxplot()+
  facet_wrap(~ionization, scales = "free")+
  scale_fill_brewer(palette = "Set1")+
  theme_minimal()+
  theme(
    axis.title.x = element_text(size = 7),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.text = element_text(size = 7),
    legend.title = element_text(size=7),
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 7)
  )


ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure2_important_slopes_boxplots.pdf",
       plot=boxplots,
       bg = "white", width = 183, height = 90, units = "mm")


```

# Logistic regressions of top predictors
```{r}


get_single_feature_roc <- 
  function(info){
    ion <- gsub("_.*", "", info)
    feature <- gsub(".*_", "", info)
    data = slopes[[ion]] %>% bind_cols() %>% mutate(y= as.numeric(factor(status))-1)
    model <- glm(y ~ data[[feature]], data = data, family = binomial) 
    data$prob <- predict(model, type = "response")
    roc_obj <- roc(data$y, data$prob)
    print(info)
    print(roc_obj)
    roc_data <- data.frame(
      TPR = roc_obj$sensitivities, # True Positive Rate 
      FPR = 1 - roc_obj$specificities # False Positive Rate
      )
    
    roc_data <- 
      pROC:::get.coords.for.ggplot(roc_obj) %>%
      as_tibble() #%>% 
    
    return(roc_data)
  }

feature_ion_combination <- c(
  paste0(rep("pos_", length(pos_features)),pos_features),
  paste0(rep("neg_", length(neg_features)),neg_features)
)

roc_single_feature <- 
  feature_ion_combination %>% 
  map(get_single_feature_roc) %>% 
  set_names(feature_ion_combination) %>% 
  bind_rows(.id = "id") %>% 
  mutate(
    Ionization = ifelse(grepl("pos", id), "Positive Ionization", "Negative Ionization"),
    Feature = case_when(
      grepl("M328.032T305.074", id)~"Hippuric Acid",
      grepl("M297.107T53.58", id)~"Tyrosyl-aspartate",
      grepl("M490.149T372.595", id)~"M490.149T372.595",
      grepl("M586.43T746.919", id)~"M586.43T746.919",
      grepl("M326.123T126.859", id)~"M326.123T126.859",
      grepl("M515.284T470.37", id)~"Dodecanedioic Acid",
      grepl("M320.993T304.505", id)~"Hippuric Acid",
      grepl("M169.087T413.776", id)~"Azelaic Acid",
      grepl("M227.983T37.03", id)~"M227.983T37.03",
      grepl("M363.18T494.785", id)~"M363.18T494.785"
    )
  )


negative <- 
  roc_single_feature %>% 
  filter(Ionization == "Negative Ionization") %>% 
  ggplot(aes(x = `1-specificity`, y = sensitivity)) + 
  geom_line() + 
  geom_abline(linetype = "dashed") + 
  labs(x = " ", y = "sensitivity (TPR)") + 
  theme_minimal()+
  facet_grid(Ionization~Feature, scales = "free") +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7))

positive <-
  roc_single_feature %>% 
  filter(Ionization == "Positive Ionization") %>% 
  ggplot(aes(x = `1-specificity`, y = sensitivity)) + 
  geom_line() + 
  geom_abline(linetype = "dashed") + 
  labs(x = "1-specificity (FPR)", y = "sensitivity (TPR)") + 
  theme_minimal()+
  facet_grid(Ionization~Feature, scales = "free") +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7))

single_feature_rocs <- cowplot::plot_grid(negative, positive, ncol = 1)  
ggsave(filename = "~/osteoporosis_metabolomics/export/single_feature_rocs.pdf",
       plot=single_feature_rocs,
       bg = "white", width = 200, height = 90, units = "mm")


```



# Figure 1 plot of individuals in PCA

```{r}
set.seed(2)
x <- ms_pos %>% select(starts_with("M"))
x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0]
y <- ms_pos %>% 
  select(-starts_with("M"))

group_counts = table(y$group) %>% sort()
groups <- sample(names(group_counts[group_counts==12]), 3)
person_filter <- y %>% filter(group %in% groups, status == "ctrl") %>% pull(cpr_enc)

y <- y %>% 
  mutate(sample_variance = ifelse(cpr_enc %in% person_filter, cpr_enc, "other"),
         sample_variance = factor(sample_variance))

indexer <- map_dbl(y$sample_variance, ~(which(.x == c(unique(person_filter), "other"))))
names_of_samples <- c("1", "2", "3", "other")
y$sample_variance <- factor(names_of_samples[indexer], levels = rev(names_of_samples))
y$indexer <- indexer

r    <- prcomp(x = x, retx = TRUE, center = T, scale = T, rank. = 6)
variance_explained <- summary(r)$importance[2,1:(6*2)]
variance_explained <- round(variance_explained, 3)*100

variable_plotlist <- list()

pd_pos <- 
  r$x %>%
  tibble::as_tibble() %>%
  dplyr::bind_cols(y %>% dplyr::select(dplyr::all_of(c("time_point","sample_variance", "indexer", "status")))) %>%
  {.}
  
plotlist <- list()

for(i in 1:(ncol(r$x)/2)) {
  
  pd_pos <- pd_pos %>% arrange(-indexer)
  
  xvar <- names(pd_pos)[2*i-1]
  yvar <- names(pd_pos)[2*i]
  p1_pos <-
    ggplot2::ggplot(pd_pos, ggplot2::aes(x=.data[[xvar]], y=.data[[yvar]], fill=sample_variance))+
    ggplot2::geom_point(shape=21, color = "white", size=2, show.legend = F) +
    ggplot2::geom_path(
      data = 
        pd_pos %>% 
        filter(sample_variance != "other") %>% 
        group_by(sample_variance) %>% 
        arrange(time_point) %>% ungroup(),
      aes(
        x = .data[[xvar]], 
        y = .data[[yvar]],
        color = sample_variance),
        show.legend = F) +#, color="black"
    ggplot2::labs(fill = "Random donor", 
                  x = paste0(xvar, " (", variance_explained[xvar], " %)"), 
                  y = paste0(yvar, " (", variance_explained[yvar], " %)"))+
    scale_color_manual(values = c("#e4211cff", "#4daf4aff", "#b250b5ff"))+
    ggplot2::scale_fill_manual(values = c("#d9d9d9ff", "#e4211cff", "#4daf4aff", "#b250b5ff"))+
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank())+
  guides(fill = guide_legend(reverse=TRUE))
  
  p2_pos <-
    ggplot2::ggplot(pd_pos, ggplot2::aes(x=.data[[xvar]], y=.data[[yvar]], fill=status))+
    ggplot2::geom_point(shape=21, color = "white", size=2, show.legend = T) +#, color="black"
    ggplot2::labs(fill = "Status", 
                  x = paste0(xvar, " (", variance_explained[xvar], " %)"), 
                  y = paste0(yvar, " (", variance_explained[yvar], " %)"))+
    ggsci::scale_fill_d3()+
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank())
  
  plotlist[[length(plotlist)+1]] <- p1_pos
}

p1 <- cowplot::plot_grid(plotlist = plotlist, nrow=1)
p1 <- cowplot::plot_grid(p1)#, legend, rel_widths = c(3, .4))
title <- cowplot::ggdraw() + cowplot::draw_label(paste("PCA based on", ncol(x), "Features"), size = 8)

final <- cowplot::plot_grid(title, p1, nrow=2, rel_heights = c(1, 10))

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/sample_variance_pos.pdf",
       plot=final,
       bg = "white", width = 183, height = 90, units = "mm")

################## RUNNING NEGATIVE SAMPLES ####################


x <- ms_neg %>% select(starts_with("M"))
x <- x[(x %>% map(is.na) %>% map_dbl(sum))==0]
y <- ms_neg %>% 
  select(-starts_with("M"))

group_counts = table(y$group) %>% sort()
groups <- sample(names(group_counts[group_counts==12]), 3)
person_filter <- y %>% filter(group %in% groups, status == "ctrl") %>% pull(cpr_enc)

y <- y %>% 
  mutate(sample_variance = ifelse(cpr_enc %in% person_filter, cpr_enc, "other"),
         sample_variance = factor(sample_variance))

indexer <- map_dbl(y$sample_variance, ~(which(.x == c(unique(person_filter), "other"))))
names_of_samples <- c("1", "2", "3", "other")
y$sample_variance <- factor(names_of_samples[indexer], levels = rev(names_of_samples))
y$indexer <- indexer

r    <- prcomp(x = x, retx = TRUE, center = T, scale = T, rank. = 6)
variance_explained <- summary(r)$importance[2,1:(6*2)]
variance_explained <- round(variance_explained, 3)*100

pd_neg <- r$x %>%
  tibble::as_tibble() %>%
  dplyr::bind_cols(y %>% dplyr::select(dplyr::all_of(c("sample_variance", "indexer", "time_point", "status")))) %>%
  {.}
  
plotlist <- list()

for(i in 1:(ncol(r$x)/2)) {
  
  pd_neg <- pd_neg %>% arrange(-indexer)
  
  xvar <- names(pd_neg)[2*i-1]
  yvar <- names(pd_neg)[2*i]
  p1_neg <-
    ggplot2::ggplot(pd_neg, ggplot2::aes(x=.data[[xvar]], y=.data[[yvar]], fill=sample_variance))+
    ggplot2::geom_point(shape=21, color = "white", size=2, show.legend = F) +
    ggplot2::geom_path(
      data = 
        pd_neg %>% 
        filter(sample_variance != "other") %>% 
        group_by(sample_variance) %>% 
        arrange(time_point) %>% ungroup(),
      aes(
        x = .data[[xvar]], 
        y = .data[[yvar]],
        color = sample_variance),
        show.legend = F) +#, color="black"
    ggplot2::labs(fill = "Random donor", 
                  x = paste0(xvar, " (", variance_explained[xvar], " %)"), 
                  y = paste0(yvar, " (", variance_explained[yvar], " %)"))+
    scale_color_manual(values = c("#e4211cff", "#4daf4aff", "#b250b5ff"))+
    ggplot2::scale_fill_manual(values = c("#d9d9d9ff", "#e4211cff", "#4daf4aff", "#b250b5ff"))+
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank())+
  guides(fill = guide_legend(reverse=TRUE))
  
  p2_neg <-
    ggplot2::ggplot(pd_neg, ggplot2::aes(x=.data[[xvar]], y=.data[[yvar]], fill=status))+
    ggplot2::geom_point(shape=21, color = "white", size=2, show.legend = T) +#, color="black"
    ggplot2::labs(fill = "Status", 
                  x = paste0(xvar, " (", variance_explained[xvar], " %)")[1], 
                  y = paste0(yvar, " (", variance_explained[yvar], " %)")
                  )+
    ggsci::scale_fill_d3()+
    ggplot2::theme_minimal() +
    ggplot2::theme(axis.title.x = element_text(size = 7),
                   axis.title.y = element_text(size = 7),
                   axis.text.x = element_text(size = 7),
                   axis.text.y = element_text(size = 7),
                   legend.text = element_text(size = 7),
                   legend.title = element_text(size=7),
                   panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank())
  
  plotlist[[length(plotlist)+1]] <- p1_neg
}

p1 <- cowplot::plot_grid(plotlist = plotlist, nrow=1)
p1 <- cowplot::plot_grid(p1)#, legend, rel_widths = c(3, .4))
title <- cowplot::ggdraw() + cowplot::draw_label(paste("PCA based on", ncol(x), "Features"), size = 8)

final <- cowplot::plot_grid(title, p1, nrow=2, rel_heights = c(1, 10))

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/sample_variance_neg.pdf",
       plot=final,
       bg = "white", width = 183, height = 90, units = "mm")
```


```{r}

library(patchwork)

pca_analysis <- (p2_pos|p2_neg)/(p1_pos|p1_neg) + plot_layout(guides = "collect")

ggsave(filename = "/data/projects/osteoporosis_metabolomics/johan/figures/figure1_pca_analysis.pdf",
       plot=pca_analysis,
       bg = "white", width = 160, height = 110, units = "mm")

```